\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-26 papers/jfp/8-engineering.tex
Work on S8

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh4 "B1" = Just $ Task $ \fetch -> do
    formula <- fetch "B1-formula"
    evalFormula fetch formula
\end{minted}
\vspace{0.5mm}

@@

\todo{AM}{Self-tracking (?):
  https://github.com/snowleopard/build/blob/master/src/Build/SelfTracking.hs}

@@

Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
system that works directly on \hs{k} and \hs{v} values. However, some build
systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
values, allowing types that are stored only in the \Shake's build information.

As one example of richer key/value types, consider the version of
\cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
to define an \emph{oracle} rule as per~\cite{mitchell2012shake} whose value is
the result of running \cmd{gcc -}\cmd{-version} and which is volatile, making the
\cmd{gcc} version something that can be depended upon. Of course, provided the
build can express volatile dependencies and supports cutoff, the version number
could equally be written to a file and used in a similar way.

@@

\Shake users have remarked that polymorphism provides a much easier expression
of concepts, e.g.~\cite{hadrian}, but it is not essential and we therefore do
not model it in this paper\footnote{See modules \cmd{Build.Multi} and
\cmd{Build.Task.Typed} in our framework for models of multiple-output and
polymorphic tasks.}.

\todo{AM}{Tasks with multiple outputs:
  https://github.com/snowleopard/build/blob/master/src/Build/Multi.hs}

\todo{AM}{Typed tasks:
  https://github.com/snowleopard/build/blob/master/src/Build/Task/Typed.hs}

>>>>

\todo{NM}{Should we add a paragraph or two giving a bit more details about how
parallelism is supported in \Shake?}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh4 "B1" = Just $ Task $ \fetch -> do
    formula <- fetch "B1-formula"
    evalFormula fetch formula
\end{minted}
\vspace{1mm}

@@

% \vspace{1mm}
% \begin{minted}[xleftmargin=10pt]{haskell}
% data Key k     = Key k   | KeyTask k
% data Value v t = Value v | ValueTask t

% fetchValue :: Functor f => (Key k -> f (Value v t)) -> k -> f v
% fetchValue fetch key = extract <$> fetch (Key key)
%   where
%     extract (Value v) = v
%     extract _ = error "Inconsistent fetch"

% -- | Fetch a task description.
% fetchValueTask :: Functor f => (Key k -> f (Value v t)) -> k -> f t
% fetchValueTask fetch key = extract <$> fetch (KeyTask key)
%   where
%     extract (ValueTask t) = t
%     extract _ = error "Inconsistent fetch"

% -- | A model for 'Monad', works beautifully and allows storing the key on the
% -- disk.
% selfTrackingM :: forall k v t. (t -> Task Monad k v) -> Tasks Monad k t -> Tasks Monad (Key k) (Value v t)
% selfTrackingM _      _     (KeyTask _) = Nothing -- Task keys are inputs
% selfTrackingM parser tasks (Key     k) = runTask <$> tasks k
%   where
%     -- Fetch the task description, parse it, and then run the obtained task
%     runTask :: Task Monad k t -> Task Monad (Key k) (Value v t)
%     runTask act = Task $ \fetch -> do
%         task <- parser <$> run act (fetchValueTask fetch)
%         Value <$> run task (fetchValue fetch)
% \end{minted}
% \vspace{1mm}

@@

Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
system that works directly on \hs{k} and \hs{v} values. However, some build
systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
values, allowing types that are stored only in the \Shake's build information.
\Shake users have remarked that polymorphism provides a much easier expression
of concepts, e.g. see~\cite{hadrian}.

As one example of richer key/value types, consider the version of
\cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
to define an \emph{oracle} rule as per~\cite{mitchell2012shake} whose value is
the result of running \cmd{gcc -}\cmd{-version} and which is volatile, making the
\cmd{gcc} version something that can be depended upon. Of course, provided the
build can express volatile dependencies and supports cutoff, the version number
could equally be written to a file and used in a similar way, albeit at the cost
of introducing additional correctness invariants, e.g. that the file
\cmd{gcc-version.txt} does indeed contains two dot-separated integers.

Generalised algebraic data types (GADTs)~\cite{spj2006gadts} can be used to
extend our models to support ``typed'' build tasks. The key idea is to replace
the monomorphic callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
corresponding values. The idea is best explained by way of an example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Version = Version { major :: Int, minor :: Int }
    deriving (Eq, Ord)

data Key a where
    File       :: FilePath -> Key String
    GccVersion :: Key Version
\end{minted}
\vspace{1mm}

\noindent
Here we extend the usual mapping from file paths to file contents with an
additional key type \hs{GccVersion} mapped to a \hs{Version}. The task
abstraction needs to be adjusted to cope with such keys (the prefix \hs{T}
stands for ``typed''):

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Fetch k f = @\std{forall}@ v. k v -> f v

newtype TTask c k v = TTask { run :: @\std{forall}@ f. c f => Fetch k f -> f v }

type TTasks c k = @\std{forall}@ v. k v -> Maybe (TTask c k v)
\end{minted}
\vspace{1mm}

\noindent
The changes compared to the definition in~\S\ref{sec-task} are minimal: (i) the
\hs{TTask} now uses a typed \hs{Fetch} callback (we define a separate type
synonym only for readability), and (ii) the type of \hs{TTasks} is now
polymorphic in \hs{v} instead of being parameterised by a concrete \hs{v}. The
example below demonstrates how \hs{fetch} can be used to retrieve dependencies
of different types: the rule \cmd{release.txt} concatenates the contents of two
\hs{File}s, while the rule \cmd{main.o} uses the numeric \hs{GccVersion} to
determine the way the \hs{source} file should be compiled.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
example :: TTasks Monad Key
example (File "release.txt") = Just $ TTask $ \fetch -> do
    readme  <- fetch (File "README")
    license <- fetch (File "LICENSE")
    return (readme ++ license)
example (File "main.o") = Just $ TTask $ \fetch -> do
    let source = "main.c"
    version <- fetch GccVersion
    if version >= Version 8 0 then compileNew source
                              else compileOld source
example _ = Nothing
\end{minted}
\vspace{1mm}

@@

% \vspace{1mm}
% \begin{minted}[xleftmargin=10pt]{haskell}
% type Partition k = k -> [k]

% multi :: Eq k => Partition k -> Tasks Applicative [k] [v] -> Tasks Applicative [k] [v]
% multi partition tasks keys
%     | k:_ <- keys, partition k == keys = tasks keys
%     | otherwise = Just $ Task $ \fetch ->
%         sequenceA [ select k <$> fetch (partition k) | k <- keys ]
%   where
%     select k = fromMaybe (error msg) . lookup k . zip (partition k)
%     msg = "Partition invariants violated"
% \end{minted}
% \vspace{1mm}
